Constraints and Opportunities for Detecting Land Surface Phenology in Drylands

### 2021-08-05 manuscript/main.tex
minor editorial edits
- define UAV in abstract
- add data and conflict of interest statements
- timeseries -> time series

%%%%%% Abstract %%%%%%
\begin{abstract}
Land surface phenology (LSP) enables global scale tracking of ecosystem processes, but its utility is limited in drylands due to low vegetation cover and resulting low annual amplitudes of vegetation indices (VIs). Drylands play an important role in global biodiversity, food security, and the carbon cycle. Thus it is important to understand the limitations in detecting and measuring dryland dynamics.

Here, using simulated data and multi-temporal UAV imagery of a desert shrubland, we explore the feasibility of detecting LSP with respect to fractional vegetation cover, plant functional types, VI uncertainty, and two different detection algorithms. Using simulated data we found that plants with distinct VI signals, such as deciduous shrubs, can require up to 60\% fractional cover to consistently detect LSP. Evergreen plants, with lower seasonal VI amplitude, require considerably higher cover and can have undetectable phenology even with 100\% vegetation cover. Our evaluation of two algorithms showed that neither performed the best in all cases. Even with adequate cover, biases in phenological metrics can still exceed 20 days, and can never be 100\% accurate due to VI uncertainty from shadows, sensor view angle, and atmospheric interference.

We showed how high-resolution UAV imagery enables LSP studies in drylands, and highlighted important scale effects driven by within canopy VI variation. With high-resolution imagery the open canopies of drylands are beneficial as they allow for straightforward identification of individual plants, enabling the tracking of phenology at the individual level. Drylands thus have the potential to become an exemplary environment for future LSP research.

@@

\subsection{Detection Methods and Analysis}
Many remote sensing algorithms have been developed in recent decades for estimating transition dates from VI timeseries, and can be categorized in four broad categories: threshold, curvature and inflection, trend, and priori curve-based approach \cite{gao-zhang2021, zeng-wardlow2020}. Here we use two approaches to evaluate LSP dynamics in drylands: a threshold and inflection (max rate of change) model. The threshold approach uses either fixed or dynamic thresholds of VI values to identify transition dates, is an established and simple methodology, and has been used to produce phenology data products from the Advanced Very High Resolution Radiometer (AVHRR), Moderate Resolution Imaging Spectroradiometer (MODIS), and the Harmonized Landsat and Sentinel-2 (HLS) data \cite{jonsson2004, gray2019, bolton2020}. The inflection approach (referred to here as the change rate method) detects the inflection points that show the quickest changes in the VI time series, thus does not depend on a priori thresholds. Curvature and inflection approaches can capture transition dates with relatively small changes in the VI timeseries, and the curvature method has been used to produce Collection 5 MODIS phenology and VIIRS phenology \cite{zhang-friedl2003, zhang-liu2018}. Both of these approaches are commonly used in LSP studies \cite{garonna2016, jeong2011, buitenwerf2015}.

@@

@@

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig5_drone_mae.png}
    \caption{The mean absolute error (MAE) of phenological estimates from 12,000 simulated pixels, with sizes from 2-16m,  in an annual UAV imagery timeseries. Phenological metrics are the day of year for start of season (SOS), end of season (EOS) and peak of season (POS). True metrics were obtained from weekly in-situ observations throughout the year.}
    \label{fig5}
\end{figure}

>>>>

%%%%%% Abstract %%%%%%
\begin{abstract}
Land surface phenology (LSP) enables global scale tracking of ecosystem processes, but its utility is limited in drylands due to low vegetation cover and resulting low annual amplitudes of vegetation indices (VIs). Drylands play an important role in global biodiversity, food security, and the carbon cycle. Thus it is important to understand the limitations in detecting and measuring dryland dynamics. Here, using simulated data and multi-temporal unmanned aerial imagery (UAV) of a desert shrubland, we explore the feasibility of detecting LSP with respect to fractional vegetation cover, plant functional types, VI uncertainty, and two different detection algorithms. Using simulated data we found that plants with distinct VI signals, such as deciduous shrubs, can require up to 60\% fractional cover to consistently detect LSP. Evergreen plants, with lower seasonal VI amplitude, require considerably higher cover and can have undetectable phenology even with 100\% vegetation cover. Our evaluation of two algorithms showed that neither performed the best in all cases. Even with adequate cover, biases in phenological metrics can still exceed 20 days, and can never be 100\% accurate due to VI uncertainty from shadows, sensor view angle, and atmospheric interference. We showed how high-resolution UAV imagery enables LSP studies in drylands, and highlighted important scale effects driven by within canopy VI variation. With high-resolution imagery the open canopies of drylands are beneficial as they allow for straightforward identification of individual plants, enabling the tracking of phenology at the individual level. Drylands thus have the potential to become an exemplary environment for future LSP research.

@@

\subsection{Detection Methods and Analysis}
Many remote sensing algorithms have been developed in recent decades for estimating transition dates from VI time series, and can be categorized in four broad categories: threshold, curvature and inflection, trend, and priori curve-based approach \cite{gao-zhang2021, zeng-wardlow2020}. Here we use two approaches to evaluate LSP dynamics in drylands: a threshold and inflection (max rate of change) model. The threshold approach uses either fixed or dynamic thresholds of VI values to identify transition dates, is an established and simple methodology, and has been used to produce phenology data products from the Advanced Very High Resolution Radiometer (AVHRR), Moderate Resolution Imaging Spectroradiometer (MODIS), and the Harmonized Landsat and Sentinel-2 (HLS) data \cite{jonsson2004, gray2019, bolton2020}. The inflection approach (referred to here as the change rate method) detects the inflection points that show the quickest changes in the VI time series, thus does not depend on a priori thresholds. Curvature and inflection approaches can capture transition dates with relatively small changes in the VI time series, and the curvature method has been used to produce Collection 5 MODIS phenology and VIIRS phenology \cite{zhang-friedl2003, zhang-liu2018}. Both of these approaches are commonly used in LSP studies \cite{garonna2016, jeong2011, buitenwerf2015}.

@@

\section*{Data Availability}
All simulated VI data, simulated pixel time series from UAV imagery, and code for reproducing this analysis, is available in the Zenodo data repository (https://doi.org/10.5281/zenodo.4777207).

@@

\section*{Conflicts of Interest}
No potential conflict of interest was reported by the authors.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig5_drone_mae.png}
    \caption{The mean absolute error (MAE) of phenological estimates from 12,000 simulated pixels, with sizes from 2-16m,  in an annual UAV imagery time series. Phenological metrics are the day of year for start of season (SOS), end of season (EOS) and peak of season (POS). True metrics were obtained from weekly in-situ observations throughout the year.}
    \label{fig5}
\end{figure}
